Division and Factorization of p-Adic Polynomials

### 2016-01-14 polynomials/muldivfact.tex
Working on todos

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
reference?)} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \nftimes \NF(Q)
\end{array}$$
where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
There exist classical algorithms for computing these two operations
whose complexity is quasi-linear with respect to the degree.

@@

\noindent
{\bf Lattice precision.}
The notion of \emph{lattice precision} was developed in 
\cite{padicprec}. It encompasses the two previous models and has
the decisive advantage of optimality. As a counterpart, it might be
very space-consuming and time-consuming for polynomials of large
degree.

@@

\noindent
We fix an integer $n$. A lattice precision datum for a polynomial of 
degree $n$ is a lattice $H$ lying in the vector space $K_{\leq n}[X]$. 
We shall sometimes denote it $O(H)$ in order to emphasize that it should 
be considered as a precision datum. The notation $P_\app(X) + O(H)$ then 
refers to any polynomial in the $W$-affine space $P_\app(X) + H$. 
Tracking lattice precision can be done using differentials as shown in
\cite[Lemma~3.4 and Proposition~3.12]{padicprec}: if $f : K_{\leq n}[X] 
\to K_{\leq m}[X]$ denotes any strictly differentiable function with
surjective differential, under mild assumption on $H$, we have:
$$f(P_\app(X)+H) = f(P_\app(X)) + f'(P_\app(X))(H)$$
where $f'(P_\app(X))$ denotes the differential of $f$ at $P_\app(X)$. 
The equality sign reflets the optimality of the method.

As already mentioned, the jagged precision model is a particular case of 
the lattice precision. Indeed, a precision of the shape $\sum_{i=0}^n 
O(\pi^{N_i}) X^i$ corresponds to the lattice generated by the elements 
$\pi^{N_i} X^i$ ($0 \leq i \leq n$). This remark is the origin of the 
notion of \emph{diffused digits of precision} introduced in 
\cite[Definition 2.3]{preclinalg}. We shall use it repeatedly in the 
sequel in order to compare the behaviour of the three aforementioned 
precision data in concrete situations.

@@

\todo{Add introduction.}
In the sequel, we use the notation $A \div B$ (resp $A \mod B$)
for the quotient (resp. the remainder) in the Euclidean division of $A$ 
by $B$.

@@

\noindent
{\bf Jagged precision.}
\todo{Write something... Explain in particular that the fast methods
(based on fast multipliction) may lead to important looses of precision.}

@@

\todo{Explain how one can use this proposition: we split the computation
into two parts (approximation and precision)...\\
Don't forget to give the complexity.}

@@

\noindent
{\bf Lattice precision.}
We now move to lattice precision. We pick $A$ and $B$ two polynomials 
of respective degree $n$ and $d$ and assume that they are known at 
precision $O(H_A)$ and $O(H_B)$ respectively:
$$A = A_\app + O(H_A)
\quad \text{and} \quad
B = B_\app + O(H_B).$$
where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
According to the results of \cite{padicprec}, in order to determine the 
precision on $A \div B$ and $A \mod B$, we need to compute the 
differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
expanding the relation:
$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
and neglecting the terms of order $\geq 2$. We get this way 
$dA = B_\app dQ + Q_\app dB + dR$
meaning that $dQ$ and $dR$ appears respectively as the quotient and
the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.

Once this has been done, the strategy is quite similar to that explained 
for Newton precision: \todo{summerize this in one sentence.}

@@

\noindent $\bullet$
in the case of lattice precision, the precision on the output is a 
lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
(The term $G_d$ corresponds to diffused digits according to 
\cite[Definition 2.3]{preclinalg}.)

@@

\todo{Add some words about the comparison between $X^5 + 2$ and
$(X+1)^5 + 2$?}

@@

A well-known theorem \todo{(Give a reference?)} asserts that each 
extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
K[X]$ corresponds to a factorization $P = AB$ where the Newton polygon 
of $A$ (resp. $B$) is given by the part of $\NP(P)$ located at the left 
(resp. the right) of $M$. Such a factorization is often called a 
\emph{slope factorization}.

@@

\noindent
\hfill
\begin{minipage}{8cm}
The lattice $H_P$ is a first order lattice at every point
of $P_\app + H_P$ in the sense of \cite[Definition~3.3]{padicprec},
\emph{i.e.} for all $P \in P_\app + H_P$:

@@

>>>>

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \cite{dwork-geratto-sullivan:Gfunctions}*{\S 1.6} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \nftimes \NF(Q)
\end{array}$$
where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
There exist classical algorithms for computing these two operations
whose complexity is quasi-linear with respect to the degree.

@@

\noindent
{\bf Lattice precision.}
The notion of \emph{lattice precision} was developed in 
\cite{caruso-roe-vaccon:14a}. It encompasses the two previous models and has
the decisive advantage of optimality. As a counterpart, it might be
very space-consuming and time-consuming for polynomials of large
degree.

@@

\noindent
We fix an integer $n$. A lattice precision datum for a polynomial of 
degree $n$ is a lattice $H$ lying in the vector space $K_{\leq n}[X]$. 
We shall sometimes denote it $O(H)$ in order to emphasize that it should 
be considered as a precision datum. The notation $P_\app(X) + O(H)$ then 
refers to any polynomial in the $W$-affine space $P_\app(X) + H$. 
Tracking lattice precision can be done using differentials as shown in
\cite{caruso-roe-vaccon:14a}*{Lemma~3.4 and Proposition~3.12}: if $f : K_{\leq n}[X] 
\to K_{\leq m}[X]$ denotes any strictly differentiable function with
surjective differential, under mild assumption on $H$, we have:
$$f(P_\app(X)+H) = f(P_\app(X)) + f'(P_\app(X))(H)$$
where $f'(P_\app(X))$ denotes the differential of $f$ at $P_\app(X)$. 
The equality sign reflets the optimality of the method.

As already mentioned, the jagged precision model is a particular case of 
the lattice precision. Indeed, a precision of the shape $\sum_{i=0}^n 
O(\pi^{N_i}) X^i$ corresponds to the lattice generated by the elements 
$\pi^{N_i} X^i$ ($0 \leq i \leq n$). This remark is the origin of the 
notion of \emph{diffused digits of precision} introduced in 
\cite{caruso-roe-vaccon:15a}*{Definition 2.3}. We shall use it repeatedly in the 
sequel in order to compare the behaviour of the three aforementioned 
precision data in concrete situations.

@@

Euclidean division provides a building block for many algorithms associated
to polynomials in one variable.  In order to analyze the precision behavior
of such algorithms, we need to first understand the precision attached to
the quotient and remainder when dividing two polynomials.
In the sequel, we use the notation $A \div B$ and $A \mod B$
for the polynomials satisfying
\begin{align*}
A &= (A \div B) \cdot B + (A \mod B) \\
\deg(A \mod B) &< \deg(B).
\end{align*}

@@

\noindent
{\bf Flat and Jagged precision.}

The simplest method for tracking the precision of a polynomial is
to record each coefficient modulo a fixed power of $p$, yet this
method suffers when applied to polynomials with small roots.
The next obvious approach is to record the precision of each
coefficient individually, a method that we will refer to as \emph{jagged}
precision.  Jagged precision is the easiest to implement in
a computer algebra system, since standard polynomial algorithms,
such as Karatsuba multiplication, can be written for generic coefficient rings.

However, this ease of implementation has a cost:
methods designed for exact rings will often have suboptimal
precision behavior, since combining intermediate expressions
into a final answer may lose precision.  Moreover, extra precision
in the middle coefficients of a polynomial, above the Newton polygon
of the remaining terms, will have no effect on any of the values
of that polynomial.  As a consequence of these deficiencies, we
do not present a detailed analysis of this precision model here.

@@

With this result in hand, we may split the computation of
Euclidean division into two pieces, first computing approximations
$Q_\app$ and $R_\app$ and separately computing $\delta Q$ and $\delta R$.
Both the approximations and the precision can be computing in time that is
quasi-linear in the degree. \todo{Is this sufficiently precision complexity analysis?}

@@

\noindent
{\bf Lattice precision.}
We now move to lattice precision. We pick $A$ and $B$ two polynomials 
of respective degree $n$ and $d$ and assume that they are known at 
precision $O(H_A)$ and $O(H_B)$ respectively:
$$A = A_\app + O(H_A)
\quad \text{and} \quad
B = B_\app + O(H_B).$$
where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
precision on $A \div B$ and $A \mod B$, we need to compute the 
differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
expanding the relation:
$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
and neglecting the terms of order $\geq 2$. We get this way 
$dA = B_\app dQ + Q_\app dB + dR$
meaning that $dQ$ and $dR$ appears respectively as the quotient and
the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.

Once this has been done, the strategy is quite similar to that explained 
for Newton precision: compute approximations and precision lattices
separately for quotient and remainder.

@@

\noindent $\bullet$
in the case of lattice precision, the precision on the output is a 
lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
(The term $G_d$ corresponds to diffused digits according to 
\cite{caruso-roe-vaccon:15a}*{Definition 2.3}.)

@@

The modulus $(X+1)^5 + 2$ shows the advantage of working
with lattice precision in intermediate computations.  While
a small amount of the precision gain is diffuse, the number of
diffuse $p$-adic digits remains constant as $n$ increases,
and the increase in precision on the actual coefficients
increases almost on pace with $X^5 + 2$.  In contrast,
non-lattice methods for $(X+1)^5 + 2$ are on par
with the other non-Eisenstein examples.

@@

A well-known theorem \cite{dwork-geratto-sullivan:Gfunctions}*{Thm. 6.1} asserts that each 
extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
K[X]$ corresponds to a factorization $P = AB$ where the Newton polygon 
of $A$ (resp. $B$) is given by the part of $\NP(P)$ located at the left 
(resp. the right) of $M$. Such a factorization is often called a 
\emph{slope factorization}.

@@

\noindent
\hfill
\begin{minipage}{8cm}
The lattice $H_P$ is a first order lattice at every point
of $P_\app + H_P$ in the sense of \cite{caruso-roe-vaccon:14a}*{Definition~3.3},
\emph{i.e.} for all $P \in P_\app + H_P$:

@@

\bibliographystyle{plain}
\bibliography{../roebib/Biblio}
